---
title: "Please change the title of the report"
author: "Please enter your name"
date: "`r Sys.Date()`"
styler: yes
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    documentclass: report
    toc_depth: 2
    toc_float: true
editor_options: 
  chunk_output_type: console
params:
    # *Please* change the 'file_name' to exactly match your data file.
    # *Please* ensure the data file is in the same directory as this code file.
    # *Please* change the 'block' to the column name that represents your block of interest
    # *Please* change the 'ref_trt' to indicate which string represents the reference group 
    # *Please* change the 'trt_interest' to indicate which strings represents the treatments of interest
    # *Please* change the 'measure' to indciate the column that represents the outcome variable  
    #*Please* change the 'y_axis_string' to  the label for the graph representing the outcome variable
    file_name: "study.csv" 
    block: "study" 
    ref_trt: "Control" 
    trt_interest: !r c("Treatment A", "Treatment B") 
    measure: "gr" 
    y_axis_string: "Growth rate" 
---


# The Purpose

The goal of this code is to enable a versatile and user-friendly R template
script designed for conducting forest plot analyses, specifically tailored to 
meet the needs of scientists. The template is intended to be sufficiently 
generic, allowing for application to continuous outcome variables to estimate 
the effect across a user-defined blocking variable.

<!-- # Pre-read instructions -->

<!-- The following are the required inputs for this code to run successfully. -->

<!-- ## Data format:  -->
<!--     1. Data in a long format with a single header (see example ...) -->
<!--     2. Minimum columns are three columns:  outcome variabe,  block,   -->
<!--         treatment (align the names) -->
<!--     3. Note:  extra columns will be ignored.  -->
<!--     4. Missing values are specified as empty cells  or NA without any quotes -->
<!--     5. The outcome variable is a continuous quantitative variable -->
<!--     6. The block variable is a variable that has a number of defined levels,  -->
<!--         e.g., donor1, donor2, donor3 -->
<!--     7. The treatment variable has a number of defined levels, e.g., ... -->


<!-- ## Workflow: 4 step process  -->
<!-- 1. *Please*, change the title and Name to generate a bespoke report -->
<!-- 2. *Please*, define the *params*, see lines 16:29 above. -->
<!--     Note in R, the format matters. Names that have additional spaces or captial -->
<!--     letters will be treated as distinct.  -->
<!-- 3. *Please*, ensure that you have copied and pasted `renv::restore()`in the R  -->
<!-- console, and hit the `Enter` button to set up the correct package environment. -->
<!-- 4. *Please*, activate the script by click on the Knit icon above.  --> 

The following are the input arguments. Please, check and make sure your have 
given the correct arguments.

```{r input-args}
str(params)
```


   
```{r global-setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## Load functions -----------------------------------------------------------
source("functions.R")
options(repos = c(CRAN = "https://cran.rstudio.com"))

## Type III sums of squares covers unbalanced design for welfare/other reasons.
options(contrasts = c("contr.sum", "contr.poly"))

## Install and load packages ------------------------------------------------
packages <- c(
    "tidyverse", "emmeans", "here", "nlme", "pander", "metaviz", "devtools",
    "metafor", "rstatix", "magrittr", "withr", "car", "broom", "broom.mixed",
    "renv", "kableExtra"
)
install_load_packages(packages)

## Colourblind-friendly colours for ggplot ----------------------------------
cbb_palette <- c(
  "#000000", "#E69F00", "#56B4E9", "#009E73",
  "#F0E442", "#0072B2", "#D55E00", "#CC79A7"
)

# Save figures in a `figures` directory --------------------------------------
if (dir.exists("figures")) unlink("figures", recursive = TRUE)
dir.create("figures")

# Save tables in a `tables` directory ----------------------------------------
if (dir.exists("tables")) unlink("tables", recursive = TRUE)
dir.create("tables")
```



# Import data

```{r import-data}
suppressWarnings(suppressMessages({
    df <- read_csv(here::here(params$file_name))
}))

df %>% slice_head(n = 5) %>% 
    kable(caption = "First five rows of the imported data", booktabs = TRUE) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
```


# Manipulate data

We split the data into relevant pairs for comparison. That is,  
* Reference and Treatment 1  
* Reference and Treatment 2  
  ...  
  ...  
  ...  
* Reference and Treatment n  


```{r manipulate-data}
# Fix variables for further analysis
dt <- df %>%
    rename_all(tolower) %>%
    as_tibble() %>%
    mutate_if(is.character, as_factor) %>%
    dplyr::rename(block_var = params$block, measure = params$measure)


cat("Structure of data")
dt %>% distinct_if(is.factor) %>% pander()


ref_trt <- params$ref_trt # reference treatment
str_glue("Reference treatment: {ref_trt}")
trt_interest <- params$trt_interest # treatments of interest 
str_glue("Treatment of interest: {trt_interest}")


# Split study into a combination of ref treatment and a treatment of interest
dt_split_trt <- map(
    trt_interest,
    ~ filter(dt, treatment %in% c(ref_trt, .))
) %>%
    set_names(trt_interest)

cat("First five observations in each `ref + treatment` data frame")
dt_split_trt %>% 
    map(~slice_sample(.x, n = 5)) %>% pander()
```


# Visualize/explore data by blocking variable

```{r raw-data-plots}
plot_raw_data <- create_boxplot(dt, params$y_axis_string) %>% 
    print()

ggsave("figures/raw-data-plot.png", plot = plot_raw_data, 
       device = "png", width = 10, height = 7)
```


# Statistical analysis

Below we fit linear regression, linear mixed effects and random effects meta-
analysis. 
The effects/means, adjusted for other variables in the models are provided, e.g., the treatment effects on tumor burden while adjusting for different 
donors. 
Also provided are comparisons between the average effect of treatments of 
interest and the reference group.   

The linear regression model was of the form,    

**measure_of_interest ~ treatment + block + block:treatment**.   

This model helps answer the questions,

- how each `treatment` affects the `measure_of_interest`,   
- how the `measure_of_interest` might vary across different `block`s, and    
- whether the effectiveness of each `treatment` changes depending on the
`block`, i.e., `block:treatment`.   

The linear mixed regression model had the formula,   

**measure_of_interest ~ treatment + (1 | block / treatment)**,   

which models both the effects of interest (`treatment`s) and account for the
variability that is not of primary interest but could influence the results,  
i.e., the random effects of `block`s and their interaction with `treatment`s. 
This approach helps ensure that the conclusions about treatment effects are as
accurate and generalizable as possible.



## Fit linear regression models (LM) and estimate0 treatment contrast  

Here we fit a fixed effect model.  This means we are estimating the average 
effect of treatment and comparing it to the unexplained variation after 
adjusting for each block studied.  This is a robust method when there are not 
many blocks.   

The first output table for each treatment returns an estimate of the average 
effect of the treatment of interest and the reference after adjusting for other variations in the model. You should look at the **emmeans**, the **SE** as a
measure of precision, and the **p.value** columns.

The second output table for each treatment returns an average estimate of the
difference in the treatment effect compared to the reference, and you should
look at the **estimate**d effect, the **SE** as a measure of precision, and the
**p.value** columns.   

```{r fit-linear-model, results = 'asis'}
results_lm <- suppressMessages(
    fit_lm_contrast_trt(dt_split_trt, trt_interest, ref_trt)
) 

process_results(results_lm)
```



## Fit linear mixed regression models and estimate treatment contrast

Here we fit a mixed effects model accounting for the block as a random effect 
which is normally distributed. This means we are estimating the average effect 
of treatment and comparing to the unexplained variation and the variation for 
block as a random effect. This gives us an estimate that uses the observed 
block behaviour to estimate the population of blocks.   

The first output table for each treatment returns an estimate of the average 
effect of the treatment of interest and the reference after adjusting for other variations in the model. You should look at the **`emmeans`**, the **SE** as a
measure of precision, and the **p.value** columns.

The second output table for each treatment returns an average estimate of the
contrast in the treatments of interest compared to the reference and you should 
look at the **`estimate`**d effect, the SE as a measure of precision, and the p.value columns.   


```{r fit-linear-mixed-model, results = 'asis'}
results_lmm <- fit_lmm_contrast_trt(dt_split_trt)
process_results(results_lmm)
```



## Diagnostic check of fitted LMM models

**Residual** is the differences between that **Predicted** from the model and
the observed value. The distribution of the residuals can help assess the 
quality of the model. If the residuals versus the predicted are wedge shaped 
then the model is predicting some readings well and others less so, and this 
would be concerning.   

In the QQ plot we would like to see the residuals randomly scattered around the 
line indicating the distribution of the residuals is close to normal.  It is not
unusual to have a few outliers at the tails. We would be concerned by a large 
systematic deviation across the dynamic range. 


```{r model-diagnostic-plots}
# Extract LMM models
lmm_models_only <- map(results_lmm, ~ .x[[1]])

# Plot model diagnostics plots: Pearson residuals
for (i in 1:length(dt_split_trt)) {
  plt <- ResVpred(
    dt_split_trt[[i]],
    lmm_models_only[[i]],
    "treatment",
    "pearson",
    trt_interest[[i]]
  ) 
  ggsave(
    paste0("figures/resid v pred-", trt_interest[[i]], " by trt", ".png"),
    plt
  )
}

for (i in 1:length(dt_split_trt)) {
  plt <- ResVpred(
    dt_split_trt[[i]],
    lmm_models_only[[i]],
    "block_var",
    "pearson",
    trt_interest[[i]]
  ) 
  ggsave(
    paste0("figures/resid v pred-", trt_interest[[i]], " by block", ".png"),
    plt
  )
}

# Plot model diagnostics plots: Q-Q plots
model_names <- names(lmm_models_only)
approach = "pearson"
plt_qq <- pmap(
    list(dt_split_trt, lmm_models_only, approach, model_names),
    ~{
        ResidualQQplot(..1, ..2, ..3, ..4)
    }
)

plt_qq <- pmap(
    list(dt_split_trt, lmm_models_only, approach, model_names),
    ~{
        png(str_glue("figures/qqplot-{..4}.png"), width = 800, height = 800)
        ResidualQQplot(..1, ..2, ..3, ..4)
        dev.off()
    }
)
```



## Collate output data from the LM and LMM models

**treatment1** is the treatment of interest, and the statistics are estimates
of its effect on the **measure of interest** holding other sources of variation
constant.

```{r lm-lmm-model-output, results = 'asis'}
pmap(
    list(lm = results_lm, lmm = results_lmm, trt_name = trt_interest),
    function(lm, lmm, trt_name) {
        # Get estimated model parameters
        lm_fit <- summary(lm$model) %>%
            tidy() %>%
            add_column(model = "lm", .before = "term")
        lmm_fit <- lmm$model %>%
            broom::tidy() %>%
            add_column(model = "lmm", .before = "effect")
        shared_cols <- intersect(names(lm_fit), names(lmm_fit))
        # Combine estimated model parameters
        bind_rows(
            lm_fit %>% select(all_of(shared_cols)), 
            lmm_fit %>% select(all_of(shared_cols))
        ) %>%
            mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
            filter(str_detect(term, "treatment1$")) %>%
            # Create and style table of estimated model parameters
            kable(
                .,
                caption = str_glue("Treatment: {trt_name}"),
                booktabs = TRUE,
            ) %>%
            kable_styling(bootstrap_options = c("striped", "hover"))
    }
)
```



## Random effects meta-analysis

Here we use a random effect meta analysis to conduct the analysis. This approach
uses summary metrics for each block. This method is equivalent to the mixed 
effect model but has been provided because for some communities it is more 
widely used.   

For each block, we have an estimate of the effect with 95% confidence interval.
The size of the the block icon is proportional to the weight of the block in 
relation to the pooled estimate. Blocks which have lower variation are given 
more weight. At the bottom of the graph is a random effect pooled estimate. 

```{r random-effects-meta-anal, results = 'asis', message=FALSE}
df1_wide_lst <- map(dt_split_trt, ~ summariseMB_wide(.x))

df1_wide_eff_est_lst <- df1_wide_lst %>%
  map(~ escalc(
    n1i = no_DP_V, n2i = no_DP_T,
    m1i = mean_V, m2i = mean_T,
    sd1i = SD_V, sd2i = SD_T,
    data = .x, measure = "MD",
    append = TRUE
  ))

r_ma_model_lst <- df1_wide_eff_est_lst %>% 
    map(~ rma(yi, vi, data = .x))

slabs <- df1_wide_eff_est_lst %>% 
    pluck(1) %>% 
    pull(block_var) %>% 
    as.character()

# Use map2 to iterate over the list and an index simultaneously
invisible(
    map2(r_ma_model_lst, trt_interest, ~{
        fpath <- here::here("figures",
                            str_glue("forest-plot-{.y}.png"))
        png(fpath, width = 800, height = 600)
        forest(.x, slab = slabs, main = .y)
        dev.off()
    })
)
forests <- map2(
    r_ma_model_lst, trt_interest, 
    ~ {
        forest(.x, slab = slabs, main = .y)
        }
)
```



## Output data from the random effects meta-analysis model

```{r ma-output, results = "asis", message=FALSE}
dt_ma <- create_ma_output(r_ma_model_lst)

dt_ma %>% imap(
    ~ {
        df <- .x %>%
            rename(parameter = name, estimate = value) %>% 
            filter(!parameter %in% c("I2", "testOfHet")) %>% 
            kable(format = "pipe",
                  caption =
                      str_glue("R.E. meta.analysis For Treatment: {.y}")) %>%
            kable_styling(bootstrap_options = c("striped", "hover"))
    }
)
```


## Assessing heterogeneity of response across the blocks

We have also provided a heterogeneity measure, $I^2$.  This provides an 
estimate of the proportion of variability in the meta-analysis that is explained 
by differences between the blocks rather than by sampling error.

$I^2$ indicates the level of heterogeneity. It ranges between 0-100%. If 
$I^2$ ≤ 50%, blocks are considered homogeneous. If $I^2$ > 50%, the 
heterogeneity is high.   

The second row provides a statistical test of heterogeneity, **testOfHet**. 
This is the Cochran's Q test, which tests whether the variability in the
observed effect sizes or outcomes are larger than would be expected based on sampling variability alone. A significant test suggests that the true
effects/outcomes are heterogeneous.

```{r model-output-summary, results = "asis", message=FALSE}
dt_ma %>% imap(
    ~ {
        df <- .x %>%
            rename(parameter = name, estimate = value) %>% 
            filter(parameter %in% c("I2", "testOfHet")) %>% 
            kable(format = "pipe",
                  caption =
                      str_glue("R.E. meta.analysis For Treatment: {.y}")) %>%
            kable_styling(bootstrap_options = c("striped", "hover"))
    }
)
```



## Variance component analysis of fixed effects model

The proportion of variance for each term in the model was estimated from the
fixed effect model.

```{r model-variance-components, eval=FALSE}
fixed_model <- lm(
    measure ~ treatment + block_var + block_var * treatment,
    data = dt,
    na.action = "na.exclude"
)
anova_output <- anova(fixed_model)
total_variance <- sum(anova_output$`Mean Sq`)
table_output <- anova_output %>% 
    as_tibble() %>% 
    mutate(`variance%` = (`Mean Sq` / total_variance) * 100) %>% 
    rename_with(~ tolower(.x), .cols = everything()) %>% 
    add_column(
        .before = "df", 
        source = c("treatment", "block", 
                    "treatment*block", "residuals")
    )

plt_vc <- ggplot(table_output, aes(x = "", y = `variance%`, fill = source)) +
    geom_bar(stat = "identity", width = 1, color = "white") +
    coord_polar("y", start = 0) +
    scale_colour_manual(values = cbb_palette, aesthetics = "colour") +
    theme_void() +
    guides(fill = guide_legend(title = "Variation source"))
ggsave("figures/variance-components.png", plt_vc)
print(plt_vc)

output_V <- dt %>%
    filter(treatment == "PBS") %>%
    group_by(block_var) %>%
    get_summary_stats(measure,
        type = "full",
        show = c("n", "mean", "median", "sd")
    )
```


